\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-30 paper/main.tex
Update model section and moved micro AUC to methods

\todo[inline]{First part is not about evaluation but optimization
  objectives, should go in the Model section:}
As the number of drug-targets are sparsely given w.r.t.\todo{no
  acronyms} to the number of both drugs and proteins considered, the
resulting training, validation and testing datasets are highly
imbalanced. As there are only $~22.000$\todo{exact number} links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function w.r.t.\todo{no
  acronym} to binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins as justified in the results section. Furthermore, we used
early stopping to detect plateaus in the training process.

\todo[inline]{This needs to be more elaborate: add the micro- vs
  macro-averages here (formulas)}
To assess each model, we computed the \textit{area under receiver
  operating characteristic}-score (AUROC), \textit{F1-score} (F1) and
\textit{Matthews correlation coefficient} (MCC) for each considered
hyperparameter configuration setting.

\todo[inline]{Remove from here:}
Avoiding validation overfitting, we eventually tested the best
performing model on the Yamanishi dataset as presented and
discussed. Furthermore, we compare our results on this very dataset
with various other cutting edge approaches to DTI-prediction.

\subsection{Model}

\todo[inline]{Introduce the notion of top-down and bottom-up:}
In order to build a method that incorporates both top-down and
bottom-up features, we first created a model for each
individually. Hence, we assemble rich molecular structure based
features for drugs from
\textit{SmilesTransformer} \citep{SmilesTransformer} and proteins from
\textit{DeepGOPlus} \citep{DeepGoPlus}. \textit{SmilesTransformer}
introduces an autoencoder, learning over the SMILES strings and thus
the molecular organization of each drug in an unsupervised manner. On
the other hand, we take advantage of the pretrained models of
\textit{DeepGOPlus}, obtaining features from the proteins amino acid
sequence, showing significant performance in the field of protein
function prediction. Thus, both embeddings seem to suitably supplement
the following ontology based representations.

@@

As we wanted to learn from the similarity of drug side effects and
protein phenotypes we opted for a deep siamese network approach, hence
learning a high-dimensional embedding emphasizing this identity by
forcing a maximal cosine similarity between these embeddings. On the
other hand we built a deep neural network for the molecular structure
based features, also benefiting from the siamese network
architecture. Therefore, the precomputed embeddings are run through a
deep neural network transformation (DNNT) which also reduces the
eventual representation size for drugs and proteins separately. An
example structure for both types of features can be found in
Figure~\ref{fig:SiameseNetwork}.

\begin{figure}[!tpb]%figure1
	\centerline{\includegraphics[width=0.35\textwidth]{figures/siamese_network.png}}
	\caption{Siamese network applied to molecular and DL2vec
          features, utilizing deep neural network transformations
          (DNNT). The similarity function $\otimes$ yields the
          similarity between both transformed embeddings e.g. by
          computing the cosine similarity.}
	\label{fig:SiameseNetwork}
\end{figure}

1
\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger
graph convolutional model, based on the protein-protein interaction
(PPI) graph. The PPI dataset is represented by a graph $G=(V,E)$,
where each protein is represented by a vertex $v\in V$, and each edge
$e\in E\subseteq V\times V$ symbolizes an interaction between two
proteins. Additionally, we introduce mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v = x(v)$, where $d$ denotes the dimensionality of the node
features.

@@

\subsubsection{Hyperparameter tuning}
To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
w.r.t. previously described metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. \todo[inline]{Specify range of
  values searched here; even better if you have the intermediate
  results and can put them here (better: in the supplement).}

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{figures/full_model.png}}
	\caption{Full DTI prediction model based on the pretrained deep neural network transformers (DNNT) for either molecular structure or ontology based features. The transformed protein representations are added to each corresponded protein as node features for the graph convolutional steps. }
	\label{fig:Fullmodel}
\end{figure}

@@

\subsection{MicroAUC}
Continuing this line, we propose the well known metric \textit{MicroAUC} for a better evaluation for this purpose. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Note that drugs and proteins can be interchanged in this formulation, which we will denote with $MicroAUC_p$ and $MicroAUC_d$, respectively. Note further that both are applicable to all three kinds of splitting schemes, but $MicroAUC_p$ and $MicroAUC_d$ are only plausible for protein split and drug split, respectively. As we want to evaluate the models performance to find all suitable drugs for each protein individually, \textit{$MicroAUC_p$} seems to be more reasonable in comparison to both $MicroAUC_d$ and $AUROC$ with respect to the previously proposed protein split cross-validation.

>>>>

To assess each model, we compute a variety of common metrics for binary classification. As the datasets are highly imbalanced, we chose the \textit{area under receiver operating characteristic}-score (AUROC) on training, validation and testing split. Therefor, we calculated the true positive rate (TPR), false positive rate (FPR) and precision score. We will further compute true positives (TP), false positives (FP), false negatives (FN) and finally true negatives (TN). 
We calculate the ROC by plotting the TPR against the FPR for various threshold settings for the model predictions. We eventually calculate the area under ROC curve utilizing trapezoidal approximations. We will refer to this measure as $MacroAUC$.

In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Note that drugs and proteins can be interchanged in this formulation, which we will denote with $MicroAUC_p$ and $MicroAUC_d$, respectively.

\subsection{Model}
Before introducing the models core components, we will clarify the previously mentioned notions of top-down and bottom-up. We will refer to features and methods, starting from the observable characteristics of either drug or protein inferring the likely molecular mechanisms as top-down. On the other hand, bottom-up refers to approaches deducing from even finer molecular properties, such as single functional groups or amino acid sequences for drugs and proteins, respectively.

In order to build a method that incorporates both top-down and
bottom-up features, we first created a model for each
individually. Hence, we assemble rich molecular structure based
features for drugs from
\textit{SmilesTransformer} \citep{SmilesTransformer} and proteins from
\textit{DeepGOPlus} \citep{DeepGoPlus}. \textit{SmilesTransformer}
introduces an autoencoder, learning over the SMILES strings and thus
the molecular organization of each drug in an unsupervised manner. On
the other hand, we take advantage of the pretrained models of
\textit{DeepGOPlus}, obtaining features from the proteins amino acid
sequence, showing significant performance in the field of protein
function prediction. Thus, both embeddings seem to suitably supplement
the following ontology based representations.

@@

\subsubsection{Siamese neural networks and modular learnable feature transformation}
As we want to learn from the similarity of drug side effects and
protein phenotypes we opted for a deep siamese network approach, hence
learning a high-dimensional embedding emphasizing this identity by
forcing a similarity between these embeddings. On the other hand we built a deep neural network for the molecular structure based features, also benefiting from the siamese network
architecture. Computing the similarity between two representations, allows for a variety of different methods.
However, we decided for the cosine similarity measure being invariant to scaling.

Therefore, the precomputed embeddings are run through a
regular, neural learnable feature transformation (LFT) network, which also reduces the
eventual representation size for drugs and proteins separately. An
example structure for both types of features can be found in
Figure~\ref{fig:SiameseNetwork}.

While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we emphasize the versatility of this approach, as both ontology and molecular feature for both drugs and proteins are reduced to similar dimensionality. This allows for a high amount of modularity and different experimental setups by plugging different kinds of features into the model. Additionally, these pretrained features can be used for a variety of other tasks. Additionally, the ontology LFT can be reused for a variety of DL2vec based features with respect to other ontologies and hypotheses. We hereby followed the results of \textit{DL2vec}, indicating that utilizing the activation function $\sigma := \mathrm{LeakyReLU}$ leads to performance increase.

\begin{figure}[!tpb]%figure1
	\centerline{\includegraphics[width=0.35\textwidth]{figures/siamese_network.png}}
	\caption{Siamese network applied to molecular and DL2vec
          features, utilizing deep learnable feature transformations
          (LFT). The similarity function $\otimes$ yields the
          similarity between both transformed embeddings e.g. by
          computing the cosine similarity.}
	\label{fig:SiameseNetwork}
\end{figure}

\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger
graph convolutional model, based on the protein-protein interaction
(PPI) graph. The PPI dataset is represented by a graph $G=(V,E)$,
where each protein is represented by a vertex $v\in V$, and each edge
$e\in E\subseteq V\times V$ symbolizes an interaction between two
proteins. Additionally, we introduce mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v = x(v)$, where $d$ denotes the dimensionality of the node
features.

@@

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparsely given with respect to the number of both drugs and proteins considered, the resulting training, validation and testing datasets are highly
imbalanced. As there are only $22.336$ links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins as justified in the results section. Furthermore, we used
early stopping to detect plateaus in the training process.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
w.r.t. previously described metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. \todo[inline]{Specify range of
  values searched here; even better if you have the intermediate
  results and can put them here (better: in the supplement).}

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{figures/full_model.png}}
	\caption{Full DTI prediction model based on the pretrained learnable feature transformations (LFT) for either molecular structure or ontology based features. The transformed protein representations are added to each corresponded protein as node features for the graph convolutional steps. }
	\label{fig:Fullmodel}
\end{figure}

@@

\subsection{Experiments}
In this section we will present the results of \name

\subsection{MicroAUC}
Continuing this line, we propose the well known metric \textit{MicroAUC} for a better evaluation for this purpose.

Note further that both are applicable to all three kinds of splitting schemes, but $MicroAUC_p$ and $MicroAUC_d$ are only plausible for protein split and drug split, respectively. As we want to evaluate the models performance to find all suitable drugs for each protein individually, \textit{$MicroAUC_p$} seems to be more reasonable in comparison to both $MicroAUC_d$ and $AUROC$ with respect to the previously proposed protein split cross-validation.
